---
title: "Housing Project - Data preparation"
output:
  html_document:
    df_print: paged
---
This file describes the workflow for getting, cleaning and formatting the data. At the start of the script you should have only this script and two data files: the "pp-complete.csv" (price paid data) and the "ukpostcodes" (longitudes and latitudes). When the script has run these files will have been merged into 288 .Rds files for easy loading into the shiny application. 


#test
## Downloads
The price paid data was downloaded from: [GOV.UK](https://www.gov.uk/government/statistical-data-sets/price-paid-data-downloads#current-month-april-2019-data) as a complete dataset with continous monthly records from 1995-current. 

The UK postcodes spatial data was downloaded from [HERE](#https://www.freemaptools.com/download-uk-postcode-lat-lng.htm) as a .csv file containing the longitude and lattitude of every listed postcode in the UK (approximately 1 GB).




## Format
The price paid data file is a .csv file that is too large (~3.7 GB) to open in excel, R or notepad. Further, there was some trouble with the data in that some of the fields for address contained commas ',' which confused the access to the file in the format of .csv. I was not able to solve this problem so my solution was to extract only the columns prior to the address columns using AWK in the command line. The dataset in its raw format contained no column names, so these were added. Command line AWK code is given below. 

```{bash}
#I cannot seem to run this bit directly from R so I use bash
awk -F ',' '{gsub(/"/,""); print $2 "," $3 "," $4 "," $5 "," $6 "," $7 }' pp-complete.csv |sed '1i price, date, code, type, oldnew, duration' > outputfile.csv

```


## Function: split by month & merge with coordinates.
The function below queries the output file from the AWK manipulation using the R package [sqldf](https://www.rdocumentation.org/packages/sqldf/versions/0.4-11) which allows the user to query databases that are too large to open with R using SQL (standard query language). 
The function has two objectives: 1 use SQL to extract yearly subsets of the data and save them as smaller .rds files.
2. merge the data with the longitude and lattitude data that we need later for mapping.

```{r}
library(sqldf)
#load spatial data
postcodes<-read.csv("ukpostcodes.csv")
postcodes$postcode<-as.character(postcodes$postcode)

#create function
prepare<-function(year){

#define the sql query to filter for year
sql<-paste("select * from file where date like '%", year, "%'", sep="")

#from sqldf package
House <- read.csv.sql("outputfile.csv", 
                          header=T,
                          eol="\n",
                          sql = sql)

#make sure R reads the dates as dates
House$date<-as.Date(House$date)

#list of sold houses - no duplicates
list <- as.list(unique(House$code))

#merge the two datasets and name columns
datamap <- subset(postcodes, postcodes$postcode %in% list, select= c("postcode","latitude",  "longitude"))  
datamap1<-merge(datamap, House, by.x="postcode",by.y="code")
names(datamap1)<-c("postcode" , "lat" , "long", "price", "date", "type", "newbuild", "duration" )

#split dataset by year and save as .Rds file
HouseList<-split(datamap1, format(datamap1$date, "%Y-%m"))
#save(HouseList, file=paste("House", year, ".Rdata", sep=""))
for(i in names(HouseList)){
  write.csv(HouseList[[i]], paste0(i,".csv"), quote = FALSE, row.names = FALSE)

}
}

#list of years in dataset
years<-c(1995:2019, 1)

#use lapply to loop through the list and apply the function
lapply(years, prepare)
```


## Function: convert to .Rds
The folder now contains one .Rds file per year with the name format "YYYY-MM.Rds". These files still contain too many points to reasonably map in leaflet. Therefore we take another step and split the data into files by month. This next function takes the .Rds file and creats 12 smaller monthly files. These files are still large, but it is possible to plot the points. Further filtering within the application will make the map faster. 

```{r}
#create list of files & remove source data and code files
files<-list.files()

files<-files[files != "DataPreparation.Rmd"];
files<-files[files != "houseforwes.R"];
files<-files[files != "ukpostcodes.csv"];
files<-files[files != "outputfile.csv"];
files<-files[files != "UKHousePrices.Rproj"];
files<-files[files != "pp-complete.csv"];

#Fuction to read the files, convert to .Rds and delete .csv
Convert<-function(file){
  name<-substr(file, 1, 7)
  Readfile<-read.csv(file)
  saveRDS(Readfile, file=paste(name,".Rds", sep=""))
  file.remove(file);
}

#Use lapply to loop over the files with our new function
lapply(files, Convert)
```



